% Section 10 — Code and data availability (mandatory GMD section).
% Every artefact is listed with a persistent, publicly resolvable
% location. Zenodo DOIs are minted from the tagged GitHub release and
% must be filled in before the final (accepted) version.

The CRAN-PM v1.0.0 software is open source under the MIT licence. All
artefacts referenced in this manuscript are publicly available.

\paragraph{Source code.}
The source repository is hosted on GitHub at
\url{https://github.com/AmmarKheder/cran-pm} (release tag
\texttt{v1.0.0}). The exact version used for the results reported here
is archived on Zenodo with persistent identifier
\texttt{10.5281/zenodo.20285633} (the \texttt{v1.0.0} GitHub release is
automatically deposited there). Continuous integration runs the unit
and integration test suite, and builds and publishes the container
image, on every release.

\paragraph{Pre-trained model weights.}
The CRAN-PM Europe v1.0 checkpoint (\texttt{cranpm-europe-v1.0.ckpt},
$\sim$1.1\,GB) is distributed as a versioned, persistent public archive
on the Hugging Face Hub at
\url{https://huggingface.co/ammarkheder/cran-pm-europe}. It is loaded
programmatically via
\texttt{CRANPMForecaster.from\_pretrained("ammarkheder/cran-pm-europe")}.

\paragraph{Container image.}
A reproducible CPU/CUDA inference environment is published to the
GitHub Container Registry at
\url{https://ghcr.io/ammarkheder/cran-pm} (tag \texttt{v1.0.0}), built
automatically from \texttt{docker/Dockerfile} by the
\texttt{Publish Docker image} GitHub Actions workflow. Training on the
AMD MI250X partition of LUMI uses the ROCm stack through Singularity
and is documented in the repository \texttt{README}.

\paragraph{Interactive demonstrator.}
A live demonstrator that performs genuine forward passes (no
pre-recorded output) on any European coordinate for the 2022 test year
is hosted as a Hugging Face Space at
\url{https://huggingface.co/spaces/ammarkheder/cran-pm}.

\paragraph{Input data.}
\begin{itemize}
\item \textbf{ERA5} reanalysis (Hersbach et al., 2020) is available from
the Copernicus Climate Data Store
(\url{https://cds.climate.copernicus.eu}); the variables and pressure
levels used by CRAN-PM are listed in Table~\ref{tab:inputs}.

\item \textbf{CAMS global atmospheric composition} (Inness et al., 2019)
is available from the Copernicus Atmosphere Data Store
(\url{https://ads.atmosphere.copernicus.eu}).

\item \textbf{GHAP daily PM\textsubscript{2.5}} satellite-derived
analysis at 1\,km (Wei et al., 2023) is available from
\url{https://weijing-rs.github.io/product.html} and mirrored on Zenodo
under DOI \texttt{10.5281/zenodo.6398971}.

\item \textbf{GMTED2010 elevation} (Danielson and Gesch, 2011) is
available from the U.S. Geological Survey at
\url{https://www.usgs.gov/coastal-changes-and-impacts/gmted2010}.

\item \textbf{EEA air-quality stations}: the verified E1a
PM\textsubscript{2.5} observations used for validation are obtained
from the European Environment Agency air-quality download service
(\url{https://eeadmz1-downloads-webapp.azurewebsites.net}); the exact
country list and quality-control filters are encoded in
\texttt{scripts/} and recorded in the GitHub release.
\end{itemize}

\paragraph{Forecasts and ready-to-run inputs.}
The model-ready, normalised pan-European input bundles for every 2022
forecast-valid date (4192\,$\times$\,6992 grid at 0.01\textdegree,
together with the precomputed CRAN-PM forecasts and the matched EEA
station observations) are deposited as a versioned, persistent public
Hugging Face dataset at
\url{https://huggingface.co/datasets/ammarkheder/cran-pm-europe-bundles}.

\paragraph{Reproducing the experiments.}
Each figure and table in this manuscript is regenerated by the scripts
in \texttt{paper/scripts/} of the source repository, which read the
deposited forecasts and station observations above; the live
demonstrator reproduces the per-tile inference end to end. Re-training
the production model from scratch requires LUMI-equivalent resources
($\sim$64 GPU-days with \texttt{configs/europe\_multiscale.yaml}); a
reduced configuration (\texttt{configs/europe\_small\_test.yaml}) trains
in $\sim$2\,h on a single MI250X GCD or A100 for smoke-testing.
