%% CRAN-PM v1.0 — model description paper for Geoscientific Model Development
%% (GMD, Copernicus). Built with copernicus.cls v7.14 (template files in
%% _copernicus_template/, copied to paper root).

\documentclass[gmd, manuscript]{copernicus}

\graphicspath{{figures/}{figures_cases/}}

\begin{document}

\title{CRAN-PM v1.0: a multi-scale Vision Transformer for high-resolution
PM$_{2.5}$ forecasting over Europe}

\Author[1,2][ammar.kheder@lut.fi]{Ammar}{Kheder}
\Author[2,3]{Wenqing}{Peng}
\Author[2,3]{Helmi}{Toropainen}
\Author[1,2]{Zhi-Song}{Liu}
\Author[1,2,3]{Michael}{Boy}

\affil[1]{LUT University, Department of Computational Engineering,
Lappeenranta, Finland}
\affil[2]{Atmospheric Modelling Centre Lahti (AMC-Lahti), Lahti, Finland}
\affil[3]{Institute for Atmospheric and Earth System Research (INAR),
University of Helsinki, Helsinki, Finland}

\runningtitle{CRAN-PM v1.0 -- multi-scale ViT for PM$_{2.5}$ forecasting}

\runningauthor{A. Kheder et al.}

\received{}
\pubdiscuss{}
\revised{}
\accepted{}
\published{}

\firstpage{1}

\maketitle

\begin{abstract}
We present \textbf{CRAN-PM v1.0}, an open-source Python package for
short-range high-resolution surface PM$_{2.5}$ forecasting over Europe.
The underlying model is a dual-branch Vision Transformer that fuses
25\,km meteorological and chemistry fields with 1\,km satellite-derived
PM$_{2.5}$ analyses through a cross-resolution attention bridge. Beyond
its scientific accuracy (RMSE\,$=\,6.85$\,$\mu$g\,m$^{-3}$ at T+1 against
2{,}971 EEA stations, 4.7--10.7\,\% lower than the best deep-learning
baseline), the package emphasises reproducibility, GPU portability and
accessibility to atmospheric scientists. CRAN-PM ships with a stable
Python API, command-line interface, training and inference workflows,
dual support for NVIDIA CUDA and AMD ROCm, container images, a
HuggingFace model repository and a Gradio web demonstrator. We document
the software architecture, the reproducibility infrastructure and three
case studies of operational interest: the March~2022 Saharan dust
intrusion, the July~2022 Iberian wildfires, and the January~2022
Polish/Slovak winter heating episode.
\end{abstract}


\introduction
\label{sec:intro}
\input{sections/01_introduction}


\section{Related software and data ecosystem}
\label{sec:related}
\input{sections/02_related}


\section{Model description}
\label{sec:model}
\input{sections/03_model}


\section{Software design}
\label{sec:software}
\input{sections/04_software}


\section{Computational performance}
\label{sec:perf}
\input{sections/05_performance}


\section{Evaluation}
\label{sec:eval}
\input{sections/06_evaluation}


\section{Case studies}
\label{sec:cases}
\input{sections/07_cases}


\section{Ablation study}
\label{sec:ablation}
\input{sections/08_ablation}


\section{Limitations and outlook}
\label{sec:limits}
\input{sections/09_limitations}


\conclusions
\label{sec:conclusions}
\input{sections/11_conclusions}


\appendix
%% Appendix A — theoretical justification of the elevation bias and
%% the wind-guided patch ordering.
\input{sections/appendix_B_theory}

\noappendix


\codedataavailability{%
\input{sections/10_availability}
}


\authorcontribution{%
AK conceived the multi-scale architecture, implemented the CRAN-PM
package, ran all experiments and wrote the manuscript. WP
contributed to the cross-resolution attention bridge and the ablation
study. HT contributed to the data ingestion pipeline and the EEA
station evaluation. ZSL contributed to the Vision Transformer design
and supervised the deep-learning aspects. MB supervised the project,
advised on scientific framing, and acquired the LUMI compute resources.
All authors read and approved the final manuscript.
}

\competinginterests{%
The authors declare that they have no competing interests.
}

\begin{acknowledgements}
This work was carried out using computational resources from LUMI
(CSC Finland) under EuroHPC project \texttt{462001140}. We thank
Samuel Ant\~ao (AMD) for ROCm portability support, the GHAP team
(W. Wei et al.) for sharing their satellite-derived PM$_{2.5}$ analyses,
and the European Environment Agency for maintaining the open EEA
monitoring archive.
\end{acknowledgements}


%% References — single bibliography compiled from the ECCV companion
%% paper's references plus GMD-specific additions (datasets, baseline
%% software, ROCm/CUDA toolchain).
\bibliographystyle{copernicus}
\bibliography{bib/eccv_refs,bib/gmd_extra}

\end{document}
